Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-02-28 Manuscript/manuscript.tex
Added more sections and standard error

Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether or not timing
systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
neccessarily suprising given that the World Track and Field Championships occurred only 
four months ago.

The contribution of this paper is to question whether or not Seiko's timing data
was accurate and thus if Devon Allen was disqualified more so because of faulty
equipment rather than reacting illegally fast.  There is little at this point
that could be done to rectify any potential mis-qualification other than to try
to prevent it from happening again.  THis matter needs to be addressed because Devon Allen's 
disqulification could be a problem that could continue to re-appear.  If athletes
are reacting fast enough that the timing system thinks that they are reacting
illegally fast, then World Athletics needs to address this and possibly change
the minimum reaction time allowed.  Currently the barrier is 0.10 seconds, any time
faster than this would result in a disqualification in the athlete who reacted too
quickly and the race is restarted \citep{Seiko-Timing}.  Since 2011, any athletes who false 
start are not given a second chance, they are removed from competition \citep{False-Start}.  
This can have massive implications as events like the Outdoor World Track and Field Championship 
only occur typically once every two years.  It is an opportunity for athletes to earn endorsement
deals and prove to sponsors such as Nike and Adidas that
they deserve a contract.  Thus the significance of a disqualification, especially in the finals 
of a major event such as the World Track and Field Championships, are very
consequential.

@@

\section{Methods and Models}
\label{sec:methods}

@@

Linear model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -468.7 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   Older Finals & -444.1 & 225.0 & 0.1496 & 0.0028 & 5.691e-05 & 2.470e-04 \\ 
   \hline
   All Pooled & -1713 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\
   \hline
   Older Pooled & -1564 & 785.4 & 0.1562 & 0.0037 & 1.280e-04 & 6.273e-04 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.1 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   Older Finals & -465.1 & 235.5 & 6.714 & 0.1674 & 0.0859 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   Older Pooled & -1688 & 846.8 & 6.426 & 0.1923 & 0.0989 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect and link=log
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.0 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
   \hline
   Older Finals & -465.0 & 235.5 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
   \hline
   Older Pooled & -1687 & 846.6 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

When determining the most effective model to use, several differnt examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear and gamma mized effects models to be built \citep{Rpkg:lme4}.A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
either a year effect, batch effect, or both are used.  Initially a linear mixed effects
model was attempted to be implemented, but due to a right hand skew in the reaction times, 
a gamma distribution seemed better equipped to model the data (The rightskew is not 
particularly suprising as it is not possible to register a reaction time below 0.0 
seconds, resulting in a lower bound but no upper bound for outliers). In the code, the 
link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
equipped to deal with high outliers.  The graph below shows the skewed outliers: ~\ref{fig:Skew}.
\begin{figure}[h]
  \centering 
  \includegraphics[scale = 0.6]{Skew}
  \label{fig:Skew}
\end{figure}

@@

@@

>>>>

Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether or not timing
systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
neccessarily suprising given that his disqualification occurred only last August.

The contribution of this paper is to question whether or not Seiko's timing data
was accurate and thus if Devon Allen was disqualified more so because of faulty
equipment rather than reacting illegally fast.  There is little at this point
that could be done to rectify any potential mis-qualification other than to try
to prevent it from happening again.  This matter needs to be addressed because Devon Allen's 
disqulification could be a problem that could continue to re-appear.  If athletes
are reacting fast enough that the timing system thinks that they are reacting
illegally fast, then World Athletics needs to address this and possibly change
the minimum reaction time allowed.  Currently the barrier is 0.10 seconds, any time
faster than this would result in a disqualification in the athlete who reacted too
quickly and the race is restarted \citep{Seiko-Timing}.  Since 2011, any athletes who false 
start are not given a second chance, they are removed from competition \citep{False-Start}.  
This can have massive implications as events like the Outdoor World Track and Field Championship 
only occur typically once every two years.  It is an opportunity for athletes to earn endorsement
deals and prove to sponsors such as Nike and Adidas that
they deserve a contract.  Thus the significance of a disqualification, especially in the finals 
of a major event such as the World Track and Field Championships, are very
consequential.

@@

@@

Linear model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -468.7 & 0.0088 & 237.4 & 0.1481 & 0.0030 & 7.762e-05 & 2.457e-04 \\
   \hline
   Older Finals & -444.1 & 0.0075 & 225.0 & 0.1496 & 0.0028 & 5.691e-05 & 2.470e-04 \\ 
   \hline
   All Pooled & -1713 & 0.131 & 859.3 & 0.1541 & 0.0040 & 1.719e-04 & 5.926e-04 \\
   \hline
   Older Pooled & -1564 & 0.113 & 785.4 & 0.1562 & 0.0037 & 1.280e-04 & 6.273e-04 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.1 & 0.3388 & 248.5 & 6.8121 & 0.1895 & 0.1148 & 0.0112 \\
   \hline
   Older Finals & -465.1 & 0.2930 & 235.5 & 6.714 & 0.1674 & 0.0859 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.3612 & 926.9 & 6.553 & 0.2269 & 0.1305 & 0.0234 \\
   \hline
   Older Pooled & -1688 & 0.3145 & 846.8 & 6.426 & 0.1923 & 0.0989 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

Gamma model with year effect and link=log
\begin{center}
  \begin{tabular}{|c | c | c | c | c | c | c | c |} 
   \hline
   Dataset & AIC & Year se & Log Likelihood & Int effect est & Int se & Int var & Resid Var \\ [0.5ex] 
   \hline\hline
   All Finals & -491.0 & 0.0504 & 248.5 & -1.914 & 0.0280 & 0.0025 & 0.0112 \\
   \hline
   Older Finals & -465.0 & 0.0439 & 235.5 & -1.901 & 0.0251 & 0.0019 & 0.0109 \\ 
   \hline
   All Pooled & -1848 & 0.0554 & 927.0 & -1.869 & 0.0348 & 0.0031 & 0.0235 \\
   \hline
   Older Pooled & -1687 & 0.0495 & 846.6 & -1.853 & 0.0305 & 0.0024 & 0.0243 \\
   \hline
  \end{tabular}
  \end{center}

\section{GLMM Gamma Model}
When determining the most effective model to use, several differnt examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear and gamma mized effects models to be built \citep{Rpkg:lme4}.A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
either a year effect, batch effect, or both are used.  Initially a linear mixed effects
model was attempted to be implemented, but due to a right hand skew in the reaction times, 
a gamma distribution seemed better equipped to model the data (The rightskew is not 
particularly suprising as it is not possible to register a reaction time below 0.0 
seconds, resulting in a lower bound but no upper bound for outliers). In the code, the 
link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
equipped to deal with high outliers.  The graph below shows the skewed outliers: ~\ref{fig:Skew}.
\begin{figure}[h]
  \centering 
  \includegraphics[scale = 0.6]{Skew}
  \label{fig:Skew}
\end{figure}

@@

In terms of the data that was used, it is worth noting that the standard error
of the intercept was higher for all 6 comparisons of the data: finals data and 
pooled data for the linear model, gamma model, gamma and log model for when
2022 was included versus when it was not.  What this shows is that the addition
of 2022 to the linear model increased the model's variation and that the addition
of 2022 resulted in a worse model.  Although this is more of a qualitative analysis,
this is further evidence that 2022 was a significant year in the role it played
in altering the gamma model.

@@

\section{2019 and National Data}
This problem of whether or not the sensor was faulty can be further explored in
two ways. First we can look at the reaction times of athletes who competed at the
World Track and Field Championship in both 2019 and 2022 and compare the times,
and secondly we can look at how athletes performed compared to how they did at
their national/qualifiers meet.  We would expect that athletes of this caliber,
who are world class athletes to not improve significantly from one year to the
next, but that is not neccessarily the results that were found.
